### 2019-11-12 Paper_shortened_pdf.Rmd
Final changes before introduction

class             : "man"
classoption       : "noextraspace"
output            : papaja::apa6_pdf
---

@@

\vspace{\baselineskip}
Emotional adjustment is defined as the absence of internalizing symptoms and emotional dysregulation [@goodmanPsychometricPropertiesStrengths2001]. It is integral to social competence, academic achievement, and mental health [@grossHandbookEmotionRegulation2013]. Ample research has demonstrated associations between children's emotional adjustment and parenting; both in terms of quantitative involvement (the amount of time parents spend caring for the child), and in terms of qualitative parenting behaviors (how do parents behave towards children). Nevertheless, the existing literature has two major shortcomings. Firstly, in assessing linkages with children’s emotional adjustment, quantitative involvement and qualitative parenting behaviors are rarely jointly investigated. Therefore, we currently do not know how the interplay between quantitative and qualitative involvement relates to emotional development. Children’s emotional adjustment might be differently impacted by parents’  control if that parent is generally highly involved with the day-to-day minutiae of childrearing, than when (s)he is generally absent. Conversely, children’s emotional adjustment might elicit different qualitative parenting behaviors from hands-on parents, who are aware of the reasons for their children’s emotional troubles, than from out-of-touch parents, who might perceive maladjustment as a problem behavior that should be sanctioned. The present study set out to address this question.

@@

Early developmental research focused on parental socialization; the notion that parents shape child development through proximal mechanisms, such as qualitative parenting behaviors [@maccobySocializationContextFamily1983; see also @vanlissaRoleFathersMothers2018a]. Qualitative parenting behaviors have been operationalized in terms of the dimensions underlying parenting styles: Responsiveness and demandingness [@maccobySocializationContextFamily1983]. Substantial evidence indicates that responsiveness, which we operationalized in terms of warm and supportive parenting, is positively associated with children's socio-emotional adjustment [@khalequePerceivedParentalAcceptanceRejection2002; @morrisRoleFamilyContext2007].

The literature linking demandingness and emotional adjustment is less conclusive [@morrisRoleFamilyContext2007]. Demandingness encompasses several parenting practices related to setting and enforcing expectations for children's behavior. Some of these might help children meet societal expectations, whereas others are overly restrictive, and interfere with children's autonomy needs [@barberCentralityControlParenting2013]. We examined two aspects of demandingness: Consistency; the extent to which consequences for transgressions are clear and enforced, and inductive reasoning; explaining the reasons for rules and punishments. Consistency has been found to correlate strongly with emotional adjustment [@brandPerceivedParentalInconsistency1990]. There is less empirical evidence for linkages between inductive reasoning and children’s emotional adjustment. Given that inductive inductive reasoning promotes the internalization of parental expectations, including display rules for emotions [@grusecImpactParentalDiscipline1994], we hypothesized that it would be positively related to emotional adjustment.

Negative parenting, which we operationalized as parental anger, is an additional qualitative parenting dimension that is negatively linked with children's emotional adjustment [@changHarshParentingRelation2003; @eisenbergParentalReactionsChildren1999]. 
Parental anger directly undermines children's emotional adjustment [@scaramellaParentingChildrenDistress2008], and models emotional dysregulation [@changHarshParentingRelation2003]. We thus hypothesized that it would be negatively related to emotional adjustment.

Fathers' qualitative involvement is associated with emotional adjustment, even after accounting for mothers' qualitative involvement [@flouriFathersBehaviorsChildren2010]. Although some research suggests that both parents influence children similarly [@lambDevelopmentSignificanceFatherChild2004; @faganShouldResearchersConceptualize2014], there is also evidence for unique contributions of mothers and fathers. For various reasons, mothers and fathers sometimes appear to fulfill somewhat specialized roles [@bogelsDoesFatherKnow2011]. In line with the theoretical notions of "attachment" and "activation" relationships, mothers tend to encourage, and respond more supportively to, emotional expression, whereas fathers structure and sanction emotion expressions to a greater extent [@cassanoInfluenceGenderParental2007; @eisenbergParentalReactionsChildren1999]. Moreover, for adolescents' emotional adjustment, maternal warmth was found to be a more salient socialization mechanism than demandingness, whereas this pattern was reversed for fathers [@vanlissaRoleFathersMothers2018a]. Similar differentiated associations were also found for younger children's psychological adjustment [@mcdowellDifferencesMothersFathers2003]. Finally, there is some evidence that emotional adjustment is more negatively impacted by mothers' negative parenting, compared to fathers [@changHarshParentingRelation2003]; perhaps because negative parenting is incongruous with the notion of the maternal "attachment" relationship. We might therefore expect emotional adjustment to be associated more strongly with maternal warmth and anger, and with paternal demanding behaviors, such as consistency and inductive reasoning.

### Parent or child effects?

A view of the family as a dynamic system, whose members reciprocally influence one another, has become commonplace [@kerrPerceivedParentingStyle2012]. Nevertheless, bi-directional effects are not routinely investigated in the childhood literature [@flouriFathersBehaviorsChildren2010]. The available evidence suggests that maladjustment elicits less parental warmth, more demandingness, and more negative parenting [@gadeyneLongitudinalRelationsParenting2004; @nepplIntergenerationalContinuityParenting2009; @lenguaBidirectionalAssociationsTemperament2005]. 
The adolescent literature provides additional support that emotional maladjustment predicts decreased parental support, and greater demandingness [@otterpohlCrossLaggedRelationsParenting2015; @vandergiessenPerceivedAutonomySupport2014; @wijsbroekDirectionEffectsPerceived2011]. Thus, parents' display punitive responses to maladjustment, including decreased support and increased negative parenting, and corrective responses, such as increased demandingness. We therefore hypothesize that emotional adjustment will positively predict parental warmth, and negatively predict consistency, inductive reasoning, and anger.

@@

Methods commonly used to examine direction of effects between parents and children have been criticized for yielding biased results if stable between-family differences are not controlled [@hamakerCritiqueCrosslaggedPanel2015]. Innovative methods provide better estimates of dynamic reciprocal processes within families, by controlling for stable differences: The recently developed random-intercept cross-lagged panel model, which is an extension of the standard cross-lagged panel model [@hamakerCritiqueCrosslaggedPanel2015], and dynamic structural equation modeling (DSEM); a multilevel generalization of this model [@asparouhovDynamicStructuralEquation2018].

Several studies have re-examined the direction of effects between parents and children using these methods, but mostly in adolescent samples. One such study found that effects of parenting on adjustment that emerged using classical methods, disappeared when stable between-family differences in parenting and adjustment were included in the model [@keijsersParentalMonitoringAdolescent2016]. Another study similarly found that most associations between maternal and paternal qualitative parenting and adolescents' emotional adjustment, were explained by between-family level correlations. The remaining predictive effects consisted primarily of child effects, not parenting effects [@vanlissaRoleFathersMothers2018a]. The authors theorized that this finding was reflective of the relative equality of parent-child relationships in adolescence. If this is correct, we would expect to find more evidence of parenting effects in the present, younger, sample.

<!-- Hier iets over de hypotheses; dat we niet weten wat er within en wat between speelt -->

@@

The amount of time parents spend with children is associated with the quality of parenting behaviors  [@sabattiniRelationMothersFathers2004]. Indeed, more involved parents tend to experience greater parenting self-efficacy, possibly reflecting the effects of experience [@gialloMothersFathersInvolvement2013a]. Furthermore, when both parents are more involved, they monitor each other's parenting behaviors to ensure that children are raised well [@mclanahanConsequencesFatherAbsence2014]. Consequently, we expect greater joint involvement to positively predict both parents' parenting quality.

With regard to differential involvement, we might also expect the more involved parent to display more desirable qualitative parenting behaviors, due to the effects of experience [@sabattiniRelationMothersFathers2004]. However, the benefit of parents monitoring each other's child-rearing should be diminished when parents are not equally involved [@mclanahanConsequencesFatherAbsence2014]. Indeed, retrospective reports indicate that an unequal childcare distribution is associated with suboptimal parenting behaviors: Increased authoritarian parenting in mothers, and increased disengaged parenting in fathers [@sabattiniRelationMothersFathers2004]. Additionally, greater differential involvement might undermine both parents' involvement quality, through its effect on the marital relationship. Childcare responsibilities are a heavy burden [@deutschPaternalParticipationChild2001]. An unfair distribution strains the interparental relationship, interferes with sensitive parenting, and undermines child well-being [@davies2006interparental]. When care is distributed relatively equally, mothers tend to be more satisfied [@pharesFathersMothersInvolvement2009], and the family system more harmonious. Consequently, balanced co-parenting is associated with greater parental warmth [@barnettInterdependenceParentingMothers2008]. We thus expect greater differential involvement to be negatively associated with both parents' parenting quality, and with children's emotional adjustment.

@@

Bioecological theories of development indicate that children require a certain amount of care, and who provides this care is, to some extent, flexible [see @bronfenbrennerEcologyHumanDevelopment1979]. This implies that joint care should be positively associated with children's emotional adjustment. Relative involvement, on the other hand, might be negatively linked to emotional adjustment, because children feel less supported when they perceive that one of their parents is relatively uninvolved, than when both parents are equally invested in their well-being [@deutschPaternalParticipationChild2001].

@@

Interactions between involvement quality and quantity have largely been overlooked in the parenting literature [@jaffeeLifeFatherBenefits2003]. One notable exception is Feinberg's coparenting model [@feinbergInternalStructureEcological2003]. Coparenting refers to the ways parents relate to each other in the parenting role. Although the coparenting literature is typically invoked to describe child-rearing in divorced families, it also holds insights for intact families. According to Feinberg, the "division of labor" in childcare, analogous to our operationalization of relative involvement, affects qualitative child-rearing behaviors and children's adjustment directly, and moderates the link between these two constructs.

There is some empirical support for interactions between involvement quality and quantity; for instance, fathers' playful and caring involvement predicted preschoolers' emotional adjustment in school, but only when accompanied by a supportive coparenting relationship [@jiaTransactionalRelationsFather2012]. In adolescence, the quantity of paternal education-related structure, an age-appropriate form of involvement, interacted with paternal warmth to predict both academic achievement and behavioral adjustment [@wangParentalInvolvementAfrican2014]. Applying this insight from the coparenting literature, we might thus expect parents' quantitative involvement to be directly associated with their qualitative parenting behaviors and children's emotional adjustment, and to moderate effects between qualitative parenting behaviors and emotional adjustment.

## Sex differences

Sex differences are a relevant factor in the interplay between parenting and emotional adjustment. Parents tend to raise boys and girls differently; expressing more demandingness with girls [@pomerantzRoleMaternalControl1998; @vanlissaRoleFathersMothers2018a], and more anger with boys [@strausCorporalPunishmentAmerican1999]. Furthermore, social learning theory posits that the same-sex parent plays a stronger role in socialization; particularly for sex-typed behavior, such as emotion display rules [@busseySocialCognitiveTheory1999]. Finally, mounting evidence suggests that parental warmth is greatest in mother-daughter relationships, and lowest in father-son relationships [see @vanlissaRoleFathersMothers2018a]. We therefore hypothesize that parents will display more warmth, consistency, and inductive reasoning towards girls, and more anger towards boys. Furthermore, we predict stronger associations between qualitative parenting and emotional adjustment between children and their same-sex parent.

@@

The present study aims to make several contributions. First, we include both mothers and fathers, in order to tease apart their common and unique roles in children's emotional adjustment development. Second, we incorporate both qualitative and quantitative measures of involvement, and examine interactions between qualitative and quantitative involvement, which have been largely overlooked in prior work. Third, we disentangle stable between-family differences from dynamic within-family predictive processes, in order to more accurately assess the direction of effects between qualitative involvement and emotional adjustment. Finally, we examined sex as a moderator of between-family differences in all study variables and within-family predictive effects.

@@

We hypothesized that, at both the within- and between-family levels, warmth, consistency, and inductive reasoning would be positively associated, and anger would be negatively associated, with emotional adjustment. At the between-family level, we examined correlations between the random intercepts of qualitative parenting behaviors and emotional adjustment (see Table \@ref(tab:tabcors)). Random intercepts reflect the average level of each variable across all waves of the study. Counter to our hypotheses, we found no significant correlations between parental warmth and children's emotional adjustment. In support of hypotheses, we found significant positive correlations between maternal and paternal consistency, and emotional adjustment. Contrary to hypotheses, we found that maternal and paternal inductive reasoning were negatively correlated with emotional adjustment. This means that, in families where parents use more inductive reasoning, children displayed poorer emotional adjustment. Finally, as hypothesized, we found significant negative correlations between maternal and paternal anger, and emotional adjustment.

@@

table_cors <- rbind(table_between_cor, table_within_cor)
write.csv(table_cors, "table_cors.csv", row.names = FALSE)
#table_cors[is.na(table_cors)] <- ""
rownames(table_cors) <- NULL
#reg_table(table_cors, caption = "Correlations between qualitative parenting and emotional adjustment", note = "Emo.: Emotional adjustment. *: <5% of posterior has opposite sign from estimate.")
reg_table(table_cors, caption = "Correlations between qualitative parenting and emotional adjustment", note = "Emo.: Emotional adjustment. *: <5% of posterior has opposite sign from estimate.")
```

We also examined within-family correlations, which are based on the residuals of the study variables, across all waves and families (Table \@ref(tab:tabcors)). In support of hypotheses, we found significant positive within-family level associations between emotional adjustment and fathers' and mothers' warmth. Notably, these associations were nearly twice as strong for fathers' warmth, as compared to mothers' warmth. In partial support of hypotheses, we found significant positive correlations of emotional adjustment with  mothers', but not fathers', consistency, as well as with fathers', but not mothers', inductive reasoning. Finally, contrary to our expectations, we found no significant within-family associations between children's emotional adjustment and maternal or paternal anger.

@@

At the within-family level, we examined over-time predictive effects between qualitative parenting and emotional adjustment. The intercepts of these random effects reflect the average effect across families. Their variances indicate whether the effect differs in strength (and sign) between families. With regard to within-families parenting effects, none of the intercepts differed significantly from zero, excepting the effect of maternal warmth on children's emotional adjustment (Table \@ref(tab:tabcrosspath)). Thus, on average, mothers' warmth had a positive significant predictive effect on children's emotional adjustment. The variances of all parenting effects were all significant, however, indicating that the effect of qualitative parenting behaviors on children's emotional adjustment varies across families. Moreover, these variances were notably larger for fathers than for mothers, indicating that fathers have a more variable effect on their children through specific parenting behaviors than mothers do.

@@

# Add letters for contrast tests

stars <- tab_unst_sections$`Defined parameters`[tab_unst_sections$`Defined parameters`$Parameter %in% c("Mother effect - child effect on mother", "Father effect - child effect on father"), ]
letter <- 1
tabrows <- list(c(2,8),
                c(3, 9))

for(i in 1:length(tabrows)){
  if(any(grepl("\\*", stars[i, ]))){
    for(c in which(grepl("\\*", stars[i, ]))){
      # c = 6
      tab[tabrows[[i]], c] <- paste0(tab[tabrows[[i]], c], letters[letter])
      letter <<- letter + 1
    }
  }
}

tab$Parameter <- gsub("^Variance", "Unstandardized variance", tab$Parameter)
reg_table(tab, caption = "Standardized between-family level intercepts and unstandardized variances of within-family predictive effects", note = "Est.: Estimate. HDPI: Highest posterior density interval. Emo.: Emotional adjustment. *: <5% of posterior has opposite sign from estimate. Matching letters indicate that parameters differed significantly, see Supplementary Table S5.")
```

With regard to child effects, in line with hypotheses, the intercept of the cross-lagged effect of emotional adjustment on maternal warmth was positive and significant (Table \@ref(tab:tabcrosspath)). Thus, when children displayed better emotional adjustment than usual in one year, their mothers displayed more warmth than usual in the next year. Contrary to hypotheses, the intercept of the effect of emotional adjustment on maternal consistency was non-significant. As hypothesized, the intercept for maternal inductive reasoning was positive. Counter to hypotheses, the intercept for maternal anger was positive, not negative. Thus, when children displayed better emotional adjustment than usual in one year, their mothers displayed more anger than usual in the next year. In partial support of hypotheses, the intercept for the child effect of adjustment on paternal warmth was positive and significant. None of the other intercepts for child effects on paternal parenting were significant. The variances of all child effects were all significant but small, indicating that the effect of children's emotional adjustment on their parents' child-rearing behaviors differs somewhat between families.

We established direction of effects by testing the significance of the difference between corresponding parent- and child effects (see Table \@ref(tab:tabcrosspath) and Supplementary Table S5). All effects that were significant in one direction, and non-significant in the opposite direction, had significant difference tests. All effects that were significant (or non-significant) in both directions had non-significant difference tests.

We examined differences between fathers and mothers the same way. <!-- NOG DOEN -->

>>>>

class             : "man"
classoption       : "noextraspace"
output            : papaja::apa6_pdf
header-includes   :
    - \raggedbottom
---

@@

\vspace{\baselineskip}
Emotional adjustment is defined as the absence of internalizing symptoms and emotional dysregulation [@goodmanPsychometricPropertiesStrengths2001]. It is integral to social competence, academic achievement, and mental health [@grossHandbookEmotionRegulation2013]. Ample research has demonstrated associations between children's emotional adjustment and parenting; both in terms of quantitative involvement (the amount of time parents spend caring for their child), and in terms of qualitative parenting behaviors (how parents behave towards their children). Nevertheless, the existing literature has two major shortcomings. Firstly, in assessing linkages with children’s emotional adjustment, quantitative involvement and qualitative parenting behaviors are rarely jointly investigated. Therefore, we currently do not know how the interplay between quantitative and qualitative involvement relates to emotional development. Children’s emotional adjustment might, for example, be differently impacted by a parent's warmth if that parent is generally highly involved with the day-to-day minutiae of childrearing, than when (s)he is generally absent. Conversely, children’s emotional adjustment might elicit different qualitative parenting behaviors from hands-on parents, who are aware of the reasons for their children’s emotional troubles, than from out-of-touch parents, who might perceive maladjustment as a problem behavior that should be sanctioned. The present study set out to address this question.

@@

Early developmental research focused on parental socialization; the notion that parents shape child development through proximal mechanisms, such as qualitative parenting behaviors [@maccobySocializationContextFamily1983; see also @vanlissaRoleFathersMothers2018a]. Qualitative parenting behaviors have been operationalized in terms of the dimensions underlying parenting styles: Responsiveness and demandingness [@maccobySocializationContextFamily1983]. Substantial evidence indicates that responsiveness, which we operationalized in terms of warm and supportive parenting, is positively associated with children's socio-emotional adjustment [@khalequePerceivedParentalAcceptanceRejection2002; @morrisRoleFamilyContext2007]. Therefore we hypothesized that parental warmth would be positively associated with children's emotional adjustment, $H_{1a}$.

The literature linking demandingness and emotional adjustment is less conclusive [@morrisRoleFamilyContext2007]. Demandingness encompasses several parenting practices related to setting and enforcing expectations for children's behavior. Some of these might help children meet societal expectations, whereas others are overly restrictive, and interfere with children's autonomy needs [@barberCentralityControlParenting2013]. We examined two aspects of demandingness: Consistency; the extent to which consequences for transgressions are clear and enforced, and inductive reasoning; explaining the reasons for rules and punishments. Consistency has been found to correlate strongly with emotional adjustment [@brandPerceivedParentalInconsistency1990]. Therefore we hypothesized that consistency would be positively associated with emotional adjustment, $H_{1b}$. There is less empirical evidence for linkages between inductive reasoning and children’s emotional adjustment. Given that inductive inductive reasoning promotes the internalization of parental expectations, including display rules for emotions [@grusecImpactParentalDiscipline1994], we hypothesized that it would also be positively related to emotional adjustment, $H_{1c}$.

Negative parenting, which we operationalized as parental anger, is an additional qualitative parenting dimension that is negatively linked with children's emotional adjustment [@changHarshParentingRelation2003; @eisenbergParentalReactionsChildren1999]. 
Parental anger expressions negatively impact children's emotional adjustment [@scaramellaParentingChildrenDistress2008], because parental anger is upsetting, and because anger reflects poor parental emotion regulation, which serves as a model for children [@changHarshParentingRelation2003]. We thus hypothesized that it would be negatively related to emotional adjustment, $H_{1d}$.

### Parent or child effects?
<!-- Hier iets doen 
Was wel echt meer control, maar rigid control oid?-->
A view of the family as a dynamic system, whose members reciprocally influence one another, has become commonplace [@kerrPerceivedParentingStyle2012]. Nevertheless, bi-directional effects are not routinely investigated in the childhood literature [@flouriFathersBehaviorsChildren2010]. Consequently, most of the research cited in support of hypotheses $H_{1a-d}$ is cross-sectional, and could be taken as support for child effects as well as parent effects. We therefore hypothesize that child effects are similar in direction, and that warmth, consistency, and inductive reasoning positively predict, and anger negatively predicts, emotional adjustment, $H_{2a-d}$. This is supported by studies that investigated child effects specifically, and found that children's maladjustment elicits less parental warmth, and more negative parenting [@gadeyneLongitudinalRelationsParenting2004]. Bi-directional positive links between parental consistency and emotional adjustment have also been reported [@lenguaBidirectionalAssociationsTemperament2005]. 
The adolescent literature provides additional evidence that emotional adjustment positively predicts parental support [@otterpohlCrossLaggedRelationsParenting2015; @vandergiessenPerceivedAutonomySupport2014], and negatively predicts negative parenting [@wijsbroekDirectionEffectsPerceived2011]. Thus, in a way, parents respond positively to emotional adjustment, and negatively to maladjustment.
<!-- Neppl 2009: Adolescents' harsh behavior towards parents predicts harsh parenting, and positive behavior predicts positive parenting -->

@@

Methods commonly used to examine direction of effects between parents and children have been criticized for yielding biased results if stable between-family differences are not controlled [@hamakerCritiqueCrosslaggedPanel2015]. Innovative methods provide better estimates of dynamic reciprocal processes within families, by controlling for stable between-family differences: The recently developed random-intercept cross-lagged panel model, which is an extension of the standard cross-lagged panel model [@hamakerCritiqueCrosslaggedPanel2015], and dynamic structural equation modeling (DSEM); a multilevel generalization of this model [@asparouhovDynamicStructuralEquation2018].

Several studies have re-examined the direction of effects between parents and children using these methods, but (to our knowledge) only in adolescent samples. One such study found that effects of parenting on adjustment that emerged using classical methods, disappeared when stable between-family differences in parenting and adjustment were included in the model [@keijsersParentalMonitoringAdolescent2016]. Another study similarly found that most associations between maternal and paternal qualitative parenting and adolescents' emotional adjustment, were explained by between-family level correlations. The remaining predictive effects consisted primarily of child effects, not parenting effects [@vanlissaRoleFathersMothers2018a]. The authors theorized that this finding was reflective of the relative equality of parent-child relationships in adolescence. If this is correct, we would expect to find more evidence of parenting effects in the present, younger, sample. These two studies also illustrate that applying these new methods changes the pattern of results in unexpected ways. Consequently, it is impossible to make specific predictions about the pattern of between-family correlations and within-family predictive effects. For now, we will assume that $H_{1a-d}$ and $H_{2a-d}$ apply at both the between- and within-family level.
<!-- Hier iets over de hypotheses; dat we niet weten wat er within en wat between speelt 
We are, however, a bit cautious about formulating more focused hypotheses about within- versus between-family effects than the one formulated above. Because there is no precedent for using the random-intercept cross-lagged panel model or its multilevel DSEM extension in the literature on parental involvement and emotional adjustment, there is no literature upon which to base such hypotheses. Our examination of between and within-family effects will therefore mainly be exploratory. 
-->

### Differences between mothers and fathers

Fathers' qualitative involvement is associated with emotional adjustment, even after accounting for mothers' qualitative involvement [@flouriFathersBehaviorsChildren2010]. Although some research suggests that both parents influence children similarly [@lambDevelopmentSignificanceFatherChild2004; @faganShouldResearchersConceptualize2014], there is also evidence for unique contributions of mothers and fathers. Some have theorized that mothers and fathers might fulfill somewhat specialized roles [@bogelsDoesFatherKnow2011]. In line with the theoretical notions of "attachment" and "activation" relationships, mothers tend to encourage and respond supportively to emotional expression to a greater extent than fathers, whereas fathers structure and sanction emotion expressions more than mothers [@cassanoInfluenceGenderParental2007; @eisenbergParentalReactionsChildren1999]. Moreover, for adolescents' emotional adjustment, maternal warmth was found to be a more salient socialization mechanism than demandingness, whereas this pattern was reversed for fathers [@vanlissaRoleFathersMothers2018a]. Similar differentiated associations were also found for younger children's psychological adjustment [@mcdowellDifferencesMothersFathers2003]. Finally, there is some evidence that emotional adjustment is more negatively impacted by mothers' negative parenting, compared to fathers [@changHarshParentingRelation2003]; perhaps because negative parenting is incongruous with the notion of the maternal "attachment" relationship. We might therefore hypothesize that emotional adjustment would be more strongly associated with maternal warmth and anger, and with paternal demanding behaviors, such as consistency and inductive reasoning, $H_{3}$.

@@

The amount of time parents spend with children is associated with the quality of parenting behaviors [@sabattiniRelationMothersFathers2004]. Indeed, more involved parents tend to experience greater parenting self-efficacy, possibly reflecting the effects of experience [@gialloMothersFathersInvolvement2013a]. Furthermore, when both parents are more involved, they monitor each other's parenting behaviors to ensure that children are raised well [@mclanahanConsequencesFatherAbsence2014]. Consequently, we hypothesize that greater joint involvement will positively attenuate both parents' desirable parenting behaviors (warmth, consistency, and inductive reasoning), and negatively attenuate parental anger, $H_{4a-d}$.

By the same reasoning, we might hypothesize that greater differential involvement would attenuate the desirable parenting behaviors of the more involved parent positively, and those of the less involved parent negatively, $H_{5a-d}$ [@sabattiniRelationMothersFathers2004]. However, there are also reasons to expect greater differential involvement to be negatively associated with both parents' parenting quality, and with children's emotional adjustment. Retrospective reports indicate that an unequal childcare distribution is associated with suboptimal parenting behaviors: Increased authoritarian parenting in mothers, and increased disengaged parenting in fathers [@sabattiniRelationMothersFathers2004]. One reason might be that parents are less able to monitor each other's child-rearing behavior when they are not equally involved [@mclanahanConsequencesFatherAbsence2014]. Additionally, differential involvement might negatively impact the marital relationship, and thereby undermine both parents' involvement quality. Childcare responsibilities are a heavy burden [@deutschPaternalParticipationChild2001]. An unfair distribution strains the interparental relationship, interferes with sensitive parenting, and undermines child well-being [@davies2006interparental]. When care is distributed relatively equally, mothers tend to be more satisfied [@pharesFathersMothersInvolvement2009], and the family system more harmonious. Consequently, balanced co-parenting is associated with greater parental warmth [@barnettInterdependenceParentingMothers2008].

@@

Bioecological theories of development indicate that children require a certain amount of care, and who provides this care is, to some extent, flexible [see @bronfenbrennerEcologyHumanDevelopment1979]. We therefore hypothesize that greater joint care should positively predict children's emotional adjustment, $H_{6a}$. By contrast, we predict that greater relative involvement will negatively predict emotional adjustment, $H_{6b}$, because children feel less supported when they perceive that one of their parents is relatively uninvolved, than when both parents are equally invested in their well-being [@deutschPaternalParticipationChild2001].

@@

Interactions between involvement quality and quantity have largely been overlooked in the parenting literature [@jaffeeLifeFatherBenefits2003]. Consequently, it is not possible to substantiate specific hypotheses for such interactions. One notable exception is Feinberg's coparenting model [@feinbergInternalStructureEcological2003]. Coparenting refers to the ways parents relate to each other in the parenting role. Although the coparenting literature is typically invoked to describe child-rearing in divorced families, it also holds insights for intact families. According to Feinberg, the "division of labor" in childcare, analogous to our operationalization of relative involvement, affects qualitative child-rearing behaviors and children's adjustment directly, and moderates the link between these two constructs. Applying this insight from the coparenting literature, we might expect parents' quantitative involvement to moderate effects between qualitative parenting behaviors and emotional adjustment. Due to the limited literature, it is not possible to formulate specific hypotheses about the nature of these moderating effects. There is, however, preliminary evidence to indicate that interactions between involvement quality and quantity are relevant. For instance, fathers' playful and caring involvement positively predicted preschoolers' emotional adjustment in school, but only when accompanied by a supportive coparenting relationship [@jiaTransactionalRelationsFather2012]. Furthermore, the quantity of paternal school-related involvement interacted with paternal warmth to positively predict adolescents' academic achievement and behavioral adjustment [@wangParentalInvolvementAfrican2014]. We thus explored the moderating effects of joint and relative quantitative involvement on the interplay between qualitative parenting behaviors and emotional adjustment.

## Child sex differences

Child sex differences are a relevant factor in the interplay between parenting and emotional adjustment. Parents tend to raise boys and girls differently; expressing more demandingness with girls [@pomerantzRoleMaternalControl1998; @vanlissaRoleFathersMothers2018a], and more anger with boys [@strausCorporalPunishmentAmerican1999]. Furthermore, social learning theory posits that the same-sex parent plays a stronger role in socialization; particularly for sex-typed behavior, such as emotion display rules [@busseySocialCognitiveTheory1999]. Finally, mounting evidence suggests that parental warmth is greatest in mother-daughter relationships, and lowest in father-son relationships [see @vanlissaRoleFathersMothers2018a]. We therefore hypothesize that parents will display more warmth, consistency, and inductive reasoning towards girls than boys, and more anger towards boys than girls, $H_{7a-d}$. Furthermore, we predict stronger associations between qualitative parenting and emotional adjustment between children and their same-sex parent, $H_{8a-d}$.

@@

The present study aims to make several contributions. First, we include both mothers and fathers, in order to tease apart their common and unique roles in children's emotional adjustment development. Second, we incorporate both qualitative and quantitative measures of involvement, and examine interactions between qualitative and quantitative involvement, which have been largely overlooked in prior work. Third, we disentangle stable between-family differences from dynamic within-family predictive processes, in order to more accurately assess the direction of effects between qualitative involvement and emotional adjustment. Finally, we examined child sex as a moderator of between-family differences in all study variables and within-family predictive effects.

@@

We hypothesized that warmth, consistency, and inductive reasoning would be positively associated, and anger would be negatively associated, with emotional adjustment. At the between-family level, we examined correlations between the random intercepts of qualitative parenting behaviors and emotional adjustment (see Table \@ref(tab:tabcors)). Random intercepts reflect the average level of each variable across all waves of the study. Counter to hypothesis $H_{1a}$, we found no significant correlations between parental warmth and children's emotional adjustment. In support of hypothesis $H_{1b}$, we found significant positive correlations between maternal and paternal consistency, and emotional adjustment. Contrary to hypothesis $H_{1c}$, we found that maternal and paternal inductive reasoning were negatively correlated with emotional adjustment. This means that, in families where parents use more inductive reasoning, children displayed poorer emotional adjustment. Finally, as hypothesized in $H_{1d}$, we found significant negative correlations between maternal and paternal anger, and emotional adjustment.

@@

table_cors <- rbind(table_between_cor, table_within_cor)
write.csv(table_cors, "table_cors.csv", row.names = FALSE)
#table_cors[is.na(table_cors)] <- ""
rownames(table_cors) <- NULL
colnames(table_cors)[1] <- "Param."
#reg_table(table_cors, caption = "Correlations between qualitative parenting and emotional adjustment", note = "Emo.: Emotional adjustment. *: <5% of posterior has opposite sign from estimate.")
reg_table(table_cors, caption = "Correlations between qualitative parenting and emotional adjustment", note = "Param.: Parameter. Emo.: Emotional adjustment. *: <5% of posterior has opposite sign from estimate.")
```

We also examined within-family correlations, which are based on the residuals of the study variables, across all waves and families (Table \@ref(tab:tabcors)). In support of hypothesis $H_{1a}$, we found significant positive within-family level associations between emotional adjustment and fathers' and mothers' warmth. In partial support of hypotheses $H_{1b,c}$, we found significant positive correlations of emotional adjustment with  mothers', but not fathers', consistency, as well as with fathers', but not mothers', inductive reasoning. Finally, contrary to hypothesis $H_{1d}$, we found no significant within-family associations between children's emotional adjustment and maternal or paternal anger. Counter to hypothesis $H_{3}$, none of the differences between mothers and fathers in the strength of either between- or within-family level correlations were significant (see Supplementary Table S5). This includes the within-family correlation between warmth and emotional adjustment, whose effect size was nearly twice as large for fathers as compared to mothers.

@@

At the within-family level, we examined over-time predictive effects between qualitative parenting and emotional adjustment. The intercepts of these random effects reflect the average standardized regression coefficient across families. The variances in Table \@ref(tab:tabcrosspath) indicate whether this coefficient differs in strength (and sign) between families. We report unstandardized variances, because the standardized variances are, by definition, set to 1. With regard to within-families parenting effects, we found no evidence for hypotheses $H_{1b-d}$, as none of the intercepts differed significantly from zero. In line with hypothesis $H_{1a}$, however, on average, a maternal warmth positively predicted children's emotional adjustment in the next wave. The standardized regression coefficient of this effect, and others, is interpreted as follows: A $.99$ standard deviation increase in children's emotional adjustment in the next wave (see Table \@ref(tab:tabcrosspath)). The variances of all parenting effects were significant, indicating that the effect of qualitative parenting behaviors on children's emotional adjustment varies across families. Moreover, these variances were all significantly larger for fathers than for mothers (see Supplementary Table S5), indicating that fathers have a more variable effect on their children through specific parenting behaviors than mothers do.

@@

# Add letters for contrast tests
letter <- 1
tab <- add_letters(list("Mother effect - child effect on mother" = c(2, 8),
                 "Father effect - child effect on father" = c(3, 9)),
            tab, tab_unst_sections)

tab <- add_letters(list("F-M Parent effects on emo." = c(2, 3),
                        "F-M Variance of parenting effect" = c(5,6),
                 "F-M Child effects on parenting" = c(8, 9),
                 "F-M Variance of child effect" = c(11, 12)),
            tab, tab_unst_sections)

tab$Parameter <- gsub("^Variance", "Unstandardized variance", tab$Parameter)
names(tab)[1] <- "Param."
reg_table(tab, caption = "Standardized between-family level intercepts and unstandardized variances of within-family predictive effects", note = "Param.: Parameter. Est.: Estimate. HDPI: Highest posterior density interval. Emo.: Emotional adjustment. *: <5% of posterior has opposite sign from estimate. Matching letters indicate that parameters differed significantly; see Supplementary Table S5 for all difference tests.")
```

With regard to child effects, in line with hypothesis $H_{2a}$, the intercept of the cross-lagged effect of emotional adjustment on maternal warmth was positive and significant (Table \@ref(tab:tabcrosspath)). Thus, when children displayed better emotional adjustment than usual in one wave, their mothers displayed more warmth than usual in the next wave. Contrary to hypothesis $H_{2b}$, the intercept of the effect of emotional adjustment on maternal consistency was non-significant. As hypothesized in $H_{2c}$, the intercept for maternal inductive reasoning was positive. Counter to hypothesis $H_{2d}$, the intercept for maternal anger was positive, not negative. Thus, when children displayed better emotional adjustment than usual in one wave, their mothers displayed more anger than usual in the next wave. In support of hypothesis $H_{2a}$, the intercept for the child effect of adjustment on paternal warmth was positive and significant. None of the other intercepts ($H_{2b-d}$)for child effects on paternal parenting were significant. The variances of all child effects were all significant but small, indicating that the effect of children's emotional adjustment on their parents' child-rearing behaviors differs somewhat between families.

We established direction of effects by testing the significance of the difference between corresponding parent- and child effects (see Table \@ref(tab:tabcrosspath) and Supplementary Table S5). All effects that were significant in one direction, and non-significant in the opposite direction, had significant difference tests. All effects that were significant (or non-significant) in both directions had non-significant difference tests. We examined differences between fathers and mothers in parent- and child effects the same way. All effects that were significant for one parent but not significant for the other parent had significant difference tests, except for the child effect of emotional adjustment on parental anger, which did not differ significantly between parents.
